Kelly Cao and Rachel Robertson Project

### 2024-05-03 products/manuscript/Manuscript.qmd
Updated figures
Final edits and cleaning

---
title: "Robertson and Cao Project"
subtitle: "Part 1:Teenage HPV Vaccination Coverage and Socioeconomic Factors"
author: Kelly Cao and Rachel Robertson
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: sentence
---

@@

# Abstract 
Human Papillomavirus (HPV) is thought to be the most common sexually transmitted disease in the United States [2]. Though HPV consists of a group of over 200 related viruses, only 12 types of HPV are considered “high-risk” and associated with the incidence of six types of cancer, including oropharyngeal, cervical, penile, vulvar, vaginal, and anal cancers [4,5]. The U.S. currently recommends HPV vaccination with Gardasil 9 to immunize against six common types of high-risk HPV infections and three HPV types associated with genital warts. It is recommended that a primary dose of the HPV vaccine be given between the ages of 11 to 12 and a secondary dose administered after 6 months to 1 year from the primary [6]. Although Healthy People 2023 has set the goal of 80% HPV vaccination coverage in the U.S., only 62.6% of U.S. teenagers are up-to-date on their HPV vaccination [7]. We are interested in examining which socioeconomic and geographic factors impact HPV vaccination completion among teenagers in the U.S. To address this question, we decided to examine 2022 survey data from the National Immunization Survey for U.S. Teenagers. This survey is distributed to both providers and households of teenagers and contains a questionnaire regarding vaccination completion for several common vaccines in the U.S. immunization schedule, including HPV vaccination status. Additionally, the survey gathers information on demographics, socioeconomic status, and geographic location of the teenage patient. Using this data, we examined how several socioeconomic and demographic factors impact the status of up-to-date HPV vaccination and determined which predictors were most important in determining the up-to-date HPV vaccination status. R version 4.2.3 with R Studio was used to clean, process, explore, and analyze the survey data. Multi-step logistic regression, principal component analysis (PCA), LASSO regression, a random forest model, and an elastic net model were employed to analyze which factors are most important for determining HPV vaccination completion. The best-performing model, the random forest model, was chosen to determine the most important factors for this prediction, while the LASSO and elastic net models were used to provide a direction of association between each predictor and the outcome of HPV vaccination completion status. It was found that state is the most important factor in determining up-to-date HPV vaccination status, followed by family income, facility, age, and race or ethnicity. A high family income bracket and non-Hispanic white ethnicity were negatively associated with HPV vaccination completion while older age groups, minority ethnic groups, and lower family income were positively associated with this outcome. The relationship between state and HPV vaccination status depends on the state, which differs by legislation, cultural norms, and funding for HPV vaccination. Our results highlight the need for consistent HPV vaccination state funding and school mandates in combination with provider education to teenage patients and their parents about the efficacy and safety of the HPV vaccination.

@@

Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings [1]. HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime [2, 3]. Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts. Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells [4]. The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes. These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases. In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers [5,6]. In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006. These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types [7]. In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 [4]. Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first [8]. In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime [9].

@@

It is generally known that those of lower income and without insurance have limited healthcare access in the U.S. [10], but it will be interesting to see if this pattern is reflected at the state level. If so, policy should be examined as it may be related to healthcare access disparities. Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 [11]. More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic [12]. Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.

@@

## Cleaning Data 
The structure of the processed data was examined.  The data was examined for missing values. No NAs were present, however missing values were present as labeled factor levels. The missing factor levels were printed, identified, and replaced with NA values. A string of 27,000 NAs were present in the outcome variable, indicating vaccination completion. All rows corresponding to the NAs for the missing outcome value were removed. Approximately 160 additional rows with NA values for facility or insurance status variables were removed. Over 16,000 rows of data remained. The year column was eliminated, as all of the data is from 2022. We saved the cleaned data as an rds and csv in the 'processed-data' folder.

@@

```{r}

#| label: Table1

@@

Table1 <- readRDS(here("results/tables/table1.rds"))

kable(Table1)

```

@@

![Figure 1. Income-poverty ratio distribution](../../results/figures/income.png)

![Figure 2. The family income distribution](../../results/figures/family.income.distribution.png)

@@

![Figure 3.Facility and vaccine completion distribution](../../results/figures/facility-vaccination.png)

A total of 79% of participants had up-to-date status for receiving 1 or more shots of HPV vaccination. U.S. states and territories had an HPV vaccination completion rate ranging from 69% to 95%, fluctuating around ~80%. There are different counts of each race/ethnicity for each factor level, so the percentage of those with up-to-date vaccine status out of each race/ethnicity level was adjusted. The Hispanic ethnicity was found to have the highest vaccine completion while the non-Hispanic white ethnicity had the lowest. It was also found that for Hispanic and Non-Hispanic black ethnicities, the income poverty ratio was higher for those who were not up-to-date in HPV vaccination status, compared to those who were. The opposite was true for mixed ethnic groups, where vaccine completion corresponded with a higher mean income-poverty ratio. When examining the overall HPV vaccination status stratified by income poverty ratio, the income poverty ratio did not differ between those who are up to date and those who are not. This may be accounted for by other trends such as that shown with ethnicity, above.

![Figure 4. Distribution of income poverty ratio stratified by race and HPV vaccination completion status](../../results/figures/raceeth-income-stratified.png)

@@

 ![Figure 5. Income poverty ratio and insurance status](../../results/figures/insurance-income-stratified.png)

@@

![Table 2. Human Papillomavirus vaccine status variables and meanings.](./images/UTDHPV-variable-table.png)

The first option, P_UTDHPV was the chosen outcome as it is the most general. It has the benefit of being able to use Age as a predictor variable and also not biasing the model towards older ages who are required to receive more than one shot to be considered up-to-date. Additionally, in April of 2022, the SAGE work group of the WHO officially recommended a single-dose HPV vaccine regime, because vaccine efficacy was found to be nearly as high with a single dose as it is with the multi-dose regime. This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability (13). In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas (14). This supports our rationale for choosing the single or more dose, up-to-date, outcome variable. The variables that remained in the initial logistic regression are listed below.

![Table 3. Potential predictor variable names and meanings.](./images/Predictor-variable-table.png)

@@

![Figure 6. Residual Plot Results](../../results/figures/residual_plot.png)

@@

![Table 4. Summary of PCA Attempt](../../results/tables/pca_summary.png)

@@

![Table 5. Summary of Initial LASSO Regression](../../results/tables/LASSO_metrics.png){height=500}

@@

![Figure 7. Top Predictors Chosen by Random Forest Model](../../results/figures/variable_importance_plot.png)

@@

![Figure 8. Top Predictors chosen by LASSO Regression Model](../../results/figures/top_lasso_predictors.png)

@@

In an attempt to produce a better-performing model compared to the LASSO regression, an elastic net regression was produced using the Tidymodels framework. The elastic model employs methods from both the LASSO regression and Ridge regression by implementing a penalty parameter and mixture parameter. The model was fit initially and then tuned for both the penalty and mixture parameters using cross-validation to maximize ROC AUC. The initial fit elastic model and tuned model both performed with an ROC AUC of 0.34. The tuned model had an accuracy and kappa measure of 0.79 and 0.06, respectively. After fitting the tuned model to testing data, the accuracy and kappa were reduced to 0.79 and 0.4 respectively. The ROC AUC for this fit increased to 0.36 and the F1 measure was 0.07. The performance of the elastic net model was very similar to the LASSO model. Neither the LASSO nor elastic net model performed as well as the random forest tree model, despite the overfitting of the random forest model. The most important predictors chosen by the elastic net model were the same as the predictors chosen by the LASSO regression. The most important predictors in determining HPV vaccination completion, from most to least important, were state (Rhode Island), facility (std, school, and teen clinics), state (Vermont), state (Mississippi), state (North Dakota), state (Iowa), state (Hawaii), housing status (unknown), insurance breaks (currently uninsured or never insured since 11 years of age), and state (Nebraska).

@@

The random forest model was chosen to predict HPV vaccination completion, as it was the best-performing model for all of the performance metrics, despite potential overfitting to the data. The other two machine learning models, the LASSO and elastic net models, may also provide insight into which predictors are most important for determining HPV vaccination completion. These models are mainly in agreement with each other and display similar variable importance rankings as the random forest model. The findings will be interpreted by examining which variables were chosen by all three models, as they may hold the most importance, followed by the predictors chosen by the random forest model.

 For all three model types that were tested, state was chosen to be the most important factor in determining HPV vaccine completion status. Both the LASSO regression and elastic net models determined that Rhode Island was the state most closely associated with HPV vaccine completion, followed by Vermont, Mississippi, North Dakota, Iowa, Hawaii, and lastly, Nebraska. For both of these models, all of these states have positive coefficients associated with the outcome of up-to-date HPV vaccination, except for Mississippi, which has a negative coefficient. This may be explained by the fact that each state determines HPV vaccination legislation, policy, and funding, for teenagers aged 11 to 13. As of 2021, four states have a law requiring vaccination for school entry, including,  Hawaii, Rhode Island, Virginia, and Washington D.C (15).  As of 2019, four states recommend HPV vaccination for school attendance, by law, including, Arkansas, Delaware, West Virginia, and Wyoming (16). As of 2019, 42 U.S. states and jurisdictions provide legislative funding to promote HPV vaccination, however, Rhode Island is the only state to implement both a school entry requirement and funding for HPV vaccination (17). Additionally, Rhode Island has the highest HPV vaccination coverage out of all U.S. states and jurisdictions, exceeding 80% coverage as of 2022 (18). The states selected as most important by the models that have legislation to fund HPV vaccination campaigns include Rhode Island, North Dakota, Mississippi, and Iowa (19). States chosen as important by the model with legislation mandating HPV vaccination are Rhode Island and Hawaii. Neither Vermont nor Nebraska have legislature funding and mandating HPV vaccination, although, both are in the top 15 states for HPV vaccination coverage in the U.S. (18,19). Conversely, Mississippi has a legislature to reimburse the cost of HPV vaccination to insurance providers but has the lowest HPV vaccination coverage in the country. This is thought to be due to misconceptions and misinformation about the HPV vaccine’s purpose and the way that HPV is transmitted (18,20). The mixture of policies and vaccination coverage rate, which is not always policy dependent, demonstrates the importance of a combination of legislation to mandate and fund HPV vaccination as well as provider and patient education to support HPV vaccination.

All three models also indicated that facility and housing status are important in determining up-to-date HPV vaccination status. According to the LASSO and elastic net models, the facility being an STD clinic, school clinic, or teen clinic, is negatively associated with HPV vaccination completion, while all other facility categories (hospitals, private facilities, mixed facilities) are positively associated with the up-to-date HPV vaccination status. This may indicate that STD clinics, school clinics, and/or teen clinics neglect to recommend or educate their teenage patients on HPV vaccination. Although school-based HPV vaccination programs have shown promise in increasing HPV vaccine uptakes internationally, these programs in the U.S. face barriers such as politicization of vaccines, patient consent processes for teens, and funding for these programs (21,22). An unknown housing status was also negatively associated with the outcome of up-to-date HPV vaccination in the LASSO and elastic net models. This may reflect that housing security indirectly impacts HPV vaccine uptake in teenagers. A cross-sectional study and analysis of house-based socioeconomic status index in the U.S. in 2020, indicated that HPV vaccination initiation was related to housing and socioeconomic status, where lower quartiles corresponded to lower rates of HPV vaccine initiation (23).

Along with the variables state and facility, the random forest model indicated family income, age, and race/ ethnicity to be in the top five most important predictors for HPV vaccination completion. The LASSO and elastic net models both indicate that family income at the highest three income brackets is negatively associated with HPV vaccination completion, while the lowest three income brackets are positively associated with HPV vaccination completion. In contrast to our findings, a CDC data brief found that the percentage of teenagers who received one or more HPV vaccine doses increased as family income moved to be greater than the federal poverty level. However, this study does not differentiate vaccination levels at the uppermost income brackets (24). A recent study analyzing National Immunization Teen Surveys from 2017 to 2021 found that 64.7% of parents in advantaged income brackets had no intention to begin HPV vaccination while 40.9% of parents in disadvantaged income brackets had no intention to begin the vaccine regime. Additionally, the advantaged group was less likely to complete HPV vaccination than the disadvantaged group. For the advantaged group, the main reason for not starting the regime was safety concerns and the main reason for not completing the regime was the belief that their child was up-to-date on HPV vaccinations (25). This highlights the importance of addressing vaccination misinformation and provider education to parents on the recommended HPV vaccination schedule.

Age was selected as the fourth most important predictor of HPV vaccination completion by the random forest model, but not selected by the LASSO or elastic net models, so the direction of association for this study is unknown. A 2024 data brief by the CDC found that the percentage of teenagers who have received at least one dose of the HPV vaccination increases with age, despite the recommended age for beginning the HPV vaccination being 11 to 12 years old in the U.S. (24). Next, Race and ethnicity were selected as the fifth most important predictor of HPV vaccination completion by the random forest model, and selected by the LASSO and elastic net models, though it was not found to be the top most important predictor by this model. The LASSO and elastic net models found that being part of the non-Hispanic ethnic group is negatively associated with the up-to-date HPV vaccination status, while all other racial and ethnic groups are positively associated with the up-to-date HPV vaccination status. Current literature provides mixed results on the association between race and ethnicity and HPV vaccination status. Several studies found that Hispanic children are less likely to have received one or more HPV vaccination doses than non-Hispanic white children, while other studies indicate that vaccine initiation is lower is non-Hispanic whites than in racial and ethnic minority groups. However, these resources support that HPV vaccine follow-through may be lower in racial and ethnic minority groups, compared to lower HPV vaccine initiation rates in non-Hispanic white teenagers (24,26,27,28). Overall, larger-scale, in-depth studies are necessary to determine the association between racial and ethnic group differences in HPV vaccination initiation and completion rates in the U.S.

# Conclusion 
The analysis studying the socioeconomic factors impacting HPV vaccination completion rates among teenagers yielded many valuable insights despite some challenges that were involved.

@@

1.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.

2.	https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.

3.	https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.

4.	HPV and Cancer. In: National Cancer Institute [Internet]. 1 Mar 2019 [cited 12 Mar 2024]. Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer

5.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058.

6.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.

7.	https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).

8.	Sciences, N. A. of, Engineering, & Medicine. (2018). Factors that affect health-care utilization. Health-Care Utilization as a Proxy in Disability Determination. https://www.ncbi.nlm.nih.gov/books/NBK500097/

>>>>

---
title: "Kelly Cao and Rachel Robertson Project"
subtitle: "Socioeconomic and geographic determinants of HPV vaccination completion for US teenagers in 2022"
author: Kelly Cao and Rachel Robertson
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
---

@@

# Abstract

Human Papillomavirus (HPV) is thought to be the most common sexually transmitted disease in the United States \[2\].
Though HPV consists of a group of over 200 related viruses, only 12 types of HPV are considered “high-risk” and associated with the incidence of six types of cancer, including oropharyngeal, cervical, penile, vulvar, vaginal, and anal cancers \[4,5\].
The U.S. currently recommends HPV vaccination with Gardasil 9 to immunize against six common types of high-risk HPV infections and three HPV types associated with genital warts.
It is recommended that a primary dose of the HPV vaccine be given between the ages of 11 to 12 and a secondary dose administered after 6 months to 1 year from the primary \[6\].
Although Healthy People 2023 has set the goal of 80% HPV vaccination coverage in the U.S., only 62.6% of U.S. teenagers are up-to-date on their HPV vaccination \[7\].
We are interested in examining which socioeconomic and geographic factors impact HPV vaccination completion among teenagers in the U.S.
To address this question, we decided to examine 2022 survey data from the National Immunization Survey for U.S.
Teenagers.
This survey is distributed to both providers and households of teenagers and contains a questionnaire regarding vaccination completion for several common vaccines in the U.S. immunization schedule, including HPV vaccination status.
Additionally, the survey gathers information on demographics, socioeconomic status, and geographic location of the teenage patient.
Using this data, we examined how several socioeconomic and demographic factors impact the status of up-to-date HPV vaccination and determined which predictors were most important in determining the up-to-date HPV vaccination status.
R version 4.2.3 with R Studio was used to clean, process, explore, and analyze the survey data.
Multi-step logistic regression, principal component analysis (PCA), LASSO regression, a random forest model, and an elastic net model were employed to analyze which factors are most important for determining HPV vaccination completion.
The best-performing model, the random forest model, was chosen to determine the most important factors for this prediction, while the LASSO and elastic net models were used to provide a direction of association between each predictor and the outcome of HPV vaccination completion status.
It was found that state is the most important factor in determining up-to-date HPV vaccination status, followed by family income, facility, age, and race or ethnicity.
A high family income bracket and non-Hispanic white ethnicity were negatively associated with HPV vaccination completion while older age groups, minority ethnic groups, and lower family income were positively associated with this outcome.
The relationship between state and HPV vaccination status depends on the state, which differs by legislation, cultural norms, and funding for HPV vaccination.
Our results highlight the need for consistent HPV vaccination state funding and school mandates in combination with provider education to teenage patients and their parents about the efficacy and safety of the HPV vaccination.

@@

Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings \[1\].
HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime \[2, 3\].
Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts.
Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells \[4\].
The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes.
These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases.
In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers \[4,5\].
In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006.
These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types \[1\].
In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 \[4\].
Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first \[6\].
In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime \[7\].

@@

It is generally known that those of lower income and without insurance have limited healthcare access in the U.S. \[8\], but it will be interesting to see if this pattern is reflected at the state level.
If so, policy should be examined as it may be related to healthcare access disparities.
Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 \[9\].
More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic \[10\].
Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.

@@

## Cleaning Data

The structure of the processed data was examined.
The data was examined for missing values.
No NAs were present, however missing values were present as labeled factor levels.
The missing factor levels were printed, identified, and replaced with NA values.
A string of 27,000 NAs were present in the outcome variable, indicating vaccination completion.
All rows corresponding to the NAs for the missing outcome value were removed.
Approximately 160 additional rows with NA values for facility or insurance status variables were removed.
Over 16,000 rows of data remained.
The year column was eliminated, as all of the data is from 2022.
We saved the cleaned data as an rds and csv in the 'processed-data' folder.

@@

```{r, warning= FALSE, message = FALSE, echo = FALSE}
#| label: Table1

@@

Table1 <- readRDS(here("results/tables/summary_table.rds"))

kable(Table1) 
```

@@

![Figure 1. Income-poverty ratio distribution for study sample of U.S. teens.](../../results/figures/income.png){width="80%"}

![Figure 2. The family income distribution for the study sample of U.S. teens.](../../results/figures/family.income.distribution.png){width="80%"}

@@

![Figure 3.Facility at which survey was conducted versus vaccine completion distribution of U.S. teens.](../../results/figures/facility-vaccination.png){width="80%"}

A total of 79% of participants had up-to-date status for receiving 1 or more shots of HPV vaccination.
U.S. states and territories had an HPV vaccination completion rate ranging from 69% to 95%, fluctuating around \~80%.
There are different counts of each race/ethnicity for each factor level, so the percentage of those with up-to-date vaccine status out of each race/ethnicity level was adjusted.
The Hispanic ethnicity was found to have the highest vaccine completion while the non-Hispanic white ethnicity had the lowest.
It was also found that for Hispanic and Non-Hispanic black ethnicities, the income poverty ratio was higher for those who were not up-to-date in HPV vaccination status, compared to those who were.
The opposite was true for mixed ethnic groups, where vaccine completion corresponded with a higher mean income-poverty ratio.
When examining the overall HPV vaccination status stratified by income poverty ratio, the income poverty ratio did not differ between those who are up to date and those who are not.
This may be accounted for by other trends such as that shown with ethnicity, above.

![Figure 4. Distribution of income poverty ratio stratified by race and HPV vaccination completion status.](../../results/figures/raceeth-income-stratified.png){width="80%"}

@@

![Figure 5. Income poverty ratio and insurance status stratified by HPV vaccination status.](../../results/figures/insurance-income-stratified.png){width="80%"}

@@

![Table 2. Human Papillomavirus vaccine status variables and meanings.](../../assets/tables/UTDHPV-variable-table.png){width="80%"}

The first option, P_UTDHPV was the chosen outcome as it is the most general.
It has the benefit of being able to use Age as a predictor variable and also not biasing the model towards older ages who are required to receive more than one shot to be considered up-to-date.
Additionally, in April of 2022, the SAGE work group of the WHO officially recommended a single-dose HPV vaccine regime, because vaccine efficacy was found to be nearly as high with a single dose as it is with the multi-dose regime.
This recommendation was given with the goal of increasing vaccine coverage in areas with lower vaccine accessibility and availability \[11\].
In September of 2023, PAHO followed suit in recommending a single-dose HPV vaccine regime for the Americas \[12\].
This supports our rationale for choosing the single or more dose, up-to-date, outcome variable.
The variables that remained in the initial logistic regression are listed below.

![Table 3. Potential predictor variable names and meanings.](../../assets/tables/Predictor-variable-table.png){width="80%"}

@@

![Figure 6. Residual Plot Results](../../results/figures/residual_plot.png){width="80%"}

@@

![Table 4. Summary of PCA Attempt](../../results/tables/pca_summary.png){width="80%"}

@@

![Table 5. Summary of Initial LASSO Regression](../../results/tables/LASSO_metrics.png){width="80%"}

@@

![Figure 7. Top Predictors Chosen by Random Forest Model](../../results/figures/variable_importance_plot.png){width="80%"}

@@

![Figure 8. Top Predictors chosen by LASSO Regression Model](../../results/figures/top_lasso_predictors.png){width="80%"}

@@

In an attempt to produce a better-performing model compared to the LASSO regression, an elastic net regression was produced using the Tidymodels framework.
The elastic model employs methods from both the LASSO regression and Ridge regression by implementing a penalty parameter and mixture parameter.
The model was fit initially and then tuned for both the penalty and mixture parameters using cross-validation to maximize ROC AUC. The initial fit elastic model and tuned model both performed with an ROC AUC of 0.34.
The tuned model had an accuracy and kappa measure of 0.79 and 0.06, respectively.
After fitting the tuned model to testing data, the accuracy and kappa were reduced to 0.79 and 0.4 respectively.
The ROC AUC for this fit increased to 0.36 and the F1 measure was 0.07.
The performance of the elastic net model was very similar to the LASSO model.
Neither the LASSO nor elastic net model performed as well as the random forest tree model, despite the overfitting of the random forest model.
The most important predictors chosen by the elastic net model were the same as the predictors chosen by the LASSO regression.
The most important predictors in determining HPV vaccination completion, from most to least important, were state (Rhode Island), facility (std, school, and teen clinics), state (Vermont), state (Mississippi), state (North Dakota), state (Iowa), state (Hawaii), housing status (unknown), insurance breaks (currently uninsured or never insured since 11 years of age), and state (Nebraska).
![Figure 8. Most Important Predictors chosen by Elastic Net Model](../../results/figures/top_en_predictors.png){width="80%"}

@@

The random forest model was chosen to predict HPV vaccination completion, as it was the best-performing model for all of the performance metrics, despite potential overfitting to the data.
The other two machine learning models, the LASSO and elastic net models, may also provide insight into which predictors are most important for determining HPV vaccination completion.
These models are mainly in agreement with each other and display similar variable importance rankings as the random forest model.
The findings will be interpreted by examining which variables were chosen by all three models, as they may hold the most importance, followed by the predictors chosen by the random forest model.
For all three model types that were tested, state was chosen to be the most important factor in determining HPV vaccine completion status.
Both the LASSO regression and elastic net models determined that Rhode Island was the state most closely associated with HPV vaccine completion, followed by Vermont, Mississippi, North Dakota, Iowa, Hawaii, and lastly, Nebraska.
For both of these models, all of these states have positive coefficients associated with the outcome of up-to-date HPV vaccination, except for Mississippi, which has a negative coefficient.
This may be explained by the fact that each state determines HPV vaccination legislation, policy, and funding, for teenagers aged 11 to 13.
As of 2021, four states have a law requiring vaccination for school entry, including, Hawaii, Rhode Island, Virginia, and Washington D.C \[13\].
As of 2019, four states recommend HPV vaccination for school attendance, by law, including, Arkansas, Delaware, West Virginia, and Wyoming \[14\].
As of 2019, 42 U.S. states and jurisdictions provide legislative funding to promote HPV vaccination, however, Rhode Island is the only state to implement both a school entry requirement and funding for HPV vaccination \[15\].
Additionally, Rhode Island has the highest HPV vaccination coverage out of all U.S. states and jurisdictions, exceeding 80% coverage as of 2022 \[16\].
The states selected as most important by the models that have legislation to fund HPV vaccination campaigns include Rhode Island, North Dakota, Mississippi, and Iowa \[14\].
States chosen as important by the model with legislation mandating HPV vaccination are Rhode Island and Hawaii.
Neither Vermont nor Nebraska have legislature funding and mandating HPV vaccination, although, both are in the top 15 states for HPV vaccination coverage in the U.S. \[17,18\].
Conversely, Mississippi has a legislature to reimburse the cost of HPV vaccination to insurance providers but has the lowest HPV vaccination coverage in the country.
This is thought to be due to misconceptions and misinformation about the HPV vaccine’s purpose and the way that HPV is transmitted \[16,18\].
The mixture of policies and vaccination coverage rate, which is not always policy dependent, demonstrates the importance of a combination of legislation to mandate and fund HPV vaccination as well as provider and patient education to support HPV vaccination.
All three models also indicated that facility and housing status are important in determining up-to-date HPV vaccination status.
According to the LASSO and elastic net models, the facility being an STD clinic, school clinic, or teen clinic, is negatively associated with HPV vaccination completion, while all other facility categories (hospitals, private facilities, mixed facilities) are positively associated with the up-to-date HPV vaccination status.
This may indicate that STD clinics, school clinics, and/or teen clinics neglect to recommend or educate their teenage patients on HPV vaccination.
Although school-based HPV vaccination programs have shown promise in increasing HPV vaccine uptakes internationally, these programs in the U.S. face barriers such as politicization of vaccines, patient consent processes for teens, and funding for these programs \[18,19\].
An unknown housing status was also negatively associated with the outcome of up-to-date HPV vaccination in the LASSO and elastic net models.
This may reflect that housing security indirectly impacts HPV vaccine uptake in teenagers.
A cross-sectional study and analysis of house-based socioeconomic status index in the U.S. in 2020, indicated that HPV vaccination initiation was related to housing and socioeconomic status, where lower quartiles corresponded to lower rates of HPV vaccine initiation \[20\].
Along with the variables state and facility, the random forest model indicated family income, age, and race/ ethnicity to be in the top five most important predictors for HPV vaccination completion.
The LASSO and elastic net models both indicate that family income at the highest three income brackets is negatively associated with HPV vaccination completion, while the lowest three income brackets are positively associated with HPV vaccination completion.
In contrast to our findings, a CDC data brief found that the percentage of teenagers who received one or more HPV vaccine doses increased as family income moved to be greater than the federal poverty level.
However, this study does not differentiate vaccination levels at the uppermost income brackets \[21\].
A recent study analyzing National Immunization Teen Surveys from 2017 to 2021 found that 64.7% of parents in advantaged income brackets had no intention to begin HPV vaccination while 40.9% of parents in disadvantaged income brackets had no intention to begin the vaccine regime.
Additionally, the advantaged group was less likely to complete HPV vaccination than the disadvantaged group.
For the advantaged group, the main reason for not starting the regime was safety concerns and the main reason for not completing the regime was the belief that their child was up-to-date on HPV vaccinations \[22\].
This highlights the importance of addressing vaccination misinformation and provider education to parents on the recommended HPV vaccination schedule.
Age was selected as the fourth most important predictor of HPV vaccination completion by the random forest model, but not selected by the LASSO or elastic net models, so the direction of association for this study is unknown.
A 2024 data brief by the CDC found that the percentage of teenagers who have received at least one dose of the HPV vaccination increases with age, despite the recommended age for beginning the HPV vaccination being 11 to 12 years old in the U.S. \[21\].
Next, Race and ethnicity were selected as the fifth most important predictor of HPV vaccination completion by the random forest model, and selected by the LASSO and elastic net models, though it was not found to be the top most important predictor by this model.
The LASSO and elastic net models found that being part of the non-Hispanic ethnic group is negatively associated with the up-to-date HPV vaccination status, while all other racial and ethnic groups are positively associated with the up-to-date HPV vaccination status.
Current literature provides mixed results on the association between race and ethnicity and HPV vaccination status.
Several studies found that Hispanic children are less likely to have received one or more HPV vaccination doses than non-Hispanic white children, while other studies indicate that vaccine initiation is lower is non-Hispanic whites than in racial and ethnic minority groups.
However, these resources support that HPV vaccine follow-through may be lower in racial and ethnic minority groups, compared to lower HPV vaccine initiation rates in non-Hispanic white teenagers \[21,23,24,25\].
Overall, larger-scale, in-depth studies are necessary to determine the association between racial and ethnic group differences in HPV vaccination initiation and completion rates in the U.S.

# Conclusion

The analysis studying the socioeconomic factors impacting HPV vaccination completion rates among teenagers yielded many valuable insights despite some challenges that were involved.

@@

1.  WHO updates recommendations on HPV vaccination schedule. (2022, December 20). World Health Organization. https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule

2.  Human papillomavirus. (2022, December 29). Office on Women’s Health. https://www.womenshealth.gov/a-z-topics/human-papillomavirus

3. Quick facts: HPV-associated cancer. Quick Facts: HPV-Associated Cancer - MN Dept. of Health. (n.d.). https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:\~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.

4.  HPV and Cancer—NCI (nciglobal,ncienterprise). (2019, March 1). [cgvArticle]. National Cancer Institute. https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer

5.  HPV Vaccine | CDC. (2023, May 22). Center for Disease Control and Prevention. https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html

6.  HPV Vaccination Recommendations | CDC. (2023, May 22). Center for Disease Control and Prevention. https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html

7.  Pingali, C., Yankey, D., Elam-Evans, L. D., Markowitz, L. E., Valier, M. R., Fredua, B., Crowe, S. J., DeSisto, C. L., Stokley, S., & Singleton, J. A. (2023). Vaccination Coverage Among Adolescents Aged 13–17 Years—National Immunization Survey–Teen, United States, 2022. MMWR. Morbidity and Mortality Weekly Report, 72. https://doi.org/10.15585/mmwr.mm7234a3
